Learning action models with minimal observability

### 2018-07-24 doc/aij18/experiments.tex
small updates added after reviewing section 6

In this section we present a set of experiments designed to evaluate different aspects of \FAMA, from its performance to the quality of the learned models. Whenever applicable, we will consider both NP-complete and PSPACE-complete scenarios, in order to draw conclusions at both levels of complexity. We will also devote one experiment to assess the suitability of the metrics proposed in this paper as evaluation metrics.

\subsection{Setup}
In order to asses the performance of our approach, we evaluate \FAMA on a wide range of domains. All tested domains are IPC domains that satisfy the \strips\ requirement~\cite{fox2003pddl2}, taken from the {\sc planning.domains} repository~\cite{muise2016planning}. Table \ref{tab:domain_features} compiles the features of the thirteen domains used in the experiments. From left to right, the columns report the name of the domain, number of actions, number of predicates, maximum arity of the actions, and maximum arity of the predicates. These are all features which affect the size of $P_\Lambda$ and, hence, have an impact on the complexity of the learning task. As can be seen in table \ref{tab:domain_features}, the selected domains range from simple ones, such as \emph{blocks} and \emph{visitall}, to complex ones, like \emph{floortile} and \emph{satellite}, with higher number of actions and predicates and with higher arity.

Following, we explain the details of our experimental setup:

@@

\item {\bf Plan traces}. For each domain, we have generated 10 plan traces with 10 actions/intermediate states each using random walks. Depending on the experiment, these traces may be used for training or testing, so more details will be provided later.

\item {\bf Planner}. The classical planner we use to solve the instances that result from our compilations is {\sc Madagascar}~\cite{rintanen2014madagascar}. We use {\sc Madagascar} for two reasons: 1) its ability to deal with planning instances populated with dead-ends, and 2) when the length of the plan traces is known (\FO action sequences or \FO state trajectories), the horizon of the solution plan is also known and can be solved as a NP-complete problem. This is because, SAT-based planners can apply the actions for programming preconditions in parallel during a single planning step (and the same for actions programming effects) because these actions do not interact. Hence, we know that the programming prefix of the solution plan can be solved in two steps, independently of the number of programming actions applied.

@@

\begin{table}[hbt!]
		\begin{center}
			\begin{tabular}{l|c|c|c|c|}	
				& \multicolumn{4}{c|}{Domain features}\\ \cline{2-5}
				Domain & \# actions & \# predicates & max action arity & max predicate arity  \\
				\hline
				Blocks & 4 & 5 & 2 & 2  \\
				Driverlog & 6 & 5 & 4 & 2  \\
				Ferry & 3 & 5 & 2 & 2  \\
				Floortile & 7 & 10 & 4 & 2  \\
				Grid & 5 & 9 & 4 & 2  \\
				Gripper & 3 & 4 & 3 & 2  \\
				Hanoi & 1 & 3 & 3 & 2  \\
				Miconic & 4 & 6 & 2 & 2  \\
				Npuzzle & 1 & 3 & 3 & 2  \\
				Parking & 4 & 5 & 3 & 2  \\
				Satellite & 5 & 8 & 4 & 2  \\
				Transport & 3 & 5 & 5 & 2  \\
				Visitall & 1 & 3 & 2 & 2  \\
				Zenotravel & 5 & 4 & 5 & 2
			\end{tabular}
		\end{center}
	\caption{\small Feature description of the 13 domains used in the experiments.}
	\label{tab:domain_features}	
\end{table}

\subsection{Impact of the size of the input knowledge}
In our first experiment, we evaluate how the size of the input knowledge, i.e., $\left|\mathcal{T}\right|$ affects \FAMA. The experiment consists in increasing the size of $\mathcal{T}$ from 1 to 10, and analyze the evolution of the computation time as well as the quality of the learned models. To keep the experiment practicable, we introduce a 1000 s timeout after which the process is killed and the learned models are assigned a score of 0 for precision and recall.
Our goals with this experiment are:
\begin{enumerate}
	\item Identify the amount of input required by \FAMA to learn sound and complete models,
	\item Evaluate the scalability of \FAMA wrt the input size which, as stated in section \ref{properties}, is one of the limiting factors of our approach.
\end{enumerate}

With this in mind we have defined two case studies:
\begin{itemize}
	\item \textbf{\FO action sequence and \PO state trajectory}: This is the typical case solved by most approaches, which corresponds to a NP-complete scenario. We are assuming a degree of observability of 10\% for the state trajectory, meaning that each literal of a state has a 10\% chance of being observed.
	\item  \textbf{\NO action sequence and \NO state trajectory}: This is a PSPACE-complete scenario where both the action sequence and state trajectory are completely empty and only initial and final states are observed, i.e., $\tau = \tup{s_0, s_m}, \forall \tau \in \mathcal{T}$.
\end{itemize}

@@

With regards to the scalability of \FAMA, figure \ref{fig:np_time} paints a very interesting picture. In this figure, we can see an exponential increase in computation time for input sizes beyond 5 traces. Until an input size of 4 traces the computation time is below 1 s, but it reaches 153 s when the input is composed of 10 traces. These results match the performance of {\sc Madagascar}, since this planner is known to struggle with plan horizons beyond 160 steps (which corresponds to 8 traces in this case).

@@

The conclusion we can draw is that learning from few input samples is both our strong point and our limitation. The NP-complete results show that the learned models are considerably sound and complete, although, to put these results into perspective, we will compare ourselves with another approach in the following section. The PSPACE-complete scenario, on the other hand, show that the learned models have less quality. Nevertheless, we argue that syntax-based metrics are not appropriate for the PSPACE-complete scenario due to the vast amount of reformulations that can appear in such under-constrained learning task and we will show proof of this in section \ref{minimal}.

\subsection{Comparison with \ARMS}
Here we compare the performance of \FAMA to that of \ARMS, one of the most well-known approaches in the action model acquisition field. \ARMS works under the assumption of plan traces with \FO action sequences and \NO state trajectories, so we will also adopt this assumption. In more detail, we define a \emph{degree of observability} $\sigma$ for the state trajectory, ranging from 0\% to 100\%, that measures the probability of observing a literal, and evaluate both approaches as $\sigma$ increases using 5 plan traces as input. Note that when $\sigma = 0$ we have a \NO state trajectory and when $\sigma=100$ we have a \FO state trajectory, all cases in-between correspond to the \PO scenario.

@@

Figures \ref{fig:comparison_precision} and \ref{fig:comparison_recall} compare \FAMA and \ARMS in terms of precision and recall. The horizontal axis represents the degree of observability, while the vertical axis show the average precision (figure \ref{fig:comparison_precision}) or recall (figure \ref{fig:comparison_recall}) computed over the 14 tested domains. Remarkably, figure \ref{fig:comparison_precision} shows that \FAMA dominates in terms of precision in all cases save for \FO state trajectories. Furthermore, in average, the models learned by our approach are 15\% to 24\% more precise than those learned by \ARMS. A similar conclusion can be drawn for recall (figure \ref{fig:comparison_recall}), where the difference is even larger, meaning that our models are more complete.

@@

The overall results indicate that \FAMA outperforms \ARMS in this particular experiment, which is one with very limited input. By no means do these results imply that our approach is better, just that it is able to learn better models when few plan traces are available. We are not able to perform experiments where far more input traces are available, since \FAMA cannot handle so much input, the same as \ARMS (or any other approach) cannot handle PSPACE-complete scenarios.

\subsection{Experiments with minimal input knowledge}
\label{minimal}
In the previous experiments we have given a general view of the performance of \FAMA under different conditions. So far, experiments have shown that \FAMA is able to learn with very small amounts of input knowledge, be it due to low observability or few training samples. In this section we want to take a closer look at the action models learned from minimal input knowledge. To that end, we will limit the input to only 2 plan traces and analyze the results under different levels of observability.

@@

Table \ref{tab:results_minimum_100_10} shows the results of the first case study. As was expected from the previous experiments, the recall scores are generally higher than the precision ones, and, in fact, the models learned for 6 of the tested domains were perfectly complete. Although precision is overall lower, it is interesting to notice that the soundness of most learned sets of positive effects is flawless. With regards to the computation time, we find times below the second in all cases save one, "floor-tile", the most complex among the tested domains.

@@

The first thing to notice in table \ref{tab:results_minimum_30_30}, which collects the result of the PSPACE case study with 30\% observability, is that the values for some domains are missing. This is the case for "floor-tile", "grid", and "parking" which not only are fairly complex domains, but are also considered \emph{puzzle-like}, a quality that is known to put a strain on the planners. Interestingly enough, "hanoi" also qualifies as a \emph{puzzle-like} domain, and we can see how this is reflected on the computation time. With regards to the quality of the models, we find that the learned models retain a similar level of soundness, while their completeness has drastically worsened with respect to the previous case study. This is specially the case for the preconditions, which have gone from 0.94 recall to 0.54. This results are difficult to explain, given that in this case study the intermediate states contain more information, but our intuition is that the sequence of actions plays a big role on the quality of the learned models or, at least, on how close they are to the GTM.

@@

Next, we analyze the PSPACE case study with \NO action sequences and state trajectories (table \ref{tab:results_minimum_0_0}). Contrary to what might be expected by looking at the previous table, we are able to find solutions for all domains. This is because, although the search is less guided, there are far more possible solutions for this learning task. This wider range of solutions is also evidenced by the quality of the learned models, which in spite of being compliant, are more removed from the original GTM. In the table we find that both precision and recall have seen their scores decreased to 0.6 and 0.5, respectively. We argue, however, that syntax-based metrics are not appropriate for scenarios with minimal observability as they cannot cope with the \emph{reformulations} that occur in these circumstances.

@@

In order to illustrate this, figure \ref{fig:macroaction} shows the PDDL encoding of the action model of the {\tt stack} operator learned from plan traces with \NO action sequences and state trajectories. This action model removes a block from on top of another block and puts it down on the table, all done in a single step. There are two main differences from the expected action model for {\tt stack}: 1) it is unstacking a block instead of stacking it, and 2) instead of the arm holding the block by the end of the action, the block ends in the table. We refer to the first difference as \emph{role swapping} and it happens when actions are removed from the trace. If no actions are present in the input traces, the name of the actions lose their utility and they are effectively anonymous actions, meaning that they can interchange their behavior with any comparable action model. The second difference is commonly known as \emph{macro-action} (in this case unstack + put-down) and it can happen when there are missing states in the input traces. Reformulated action models are perfectly valid and can be used to solve planning tasks. For instance, any blocks-world problem can be solved by unstacking all the blocks to the table (unstack + put-down) and then stacking them to meet the goal condition (pick-up + stack), so reformulated action models like the one in the figure can be used to solve common planning tasks.

As we have shown, the \NO/\NO case study meets all the criteria needed for reformulations to happen, and this is the reason why scenarios such as this are better evaluated using semantic-based metrics.

@@

@@

Following, we take a look at table \ref{tab:metric_comparison_0_0} that details the results of the \NO/\NO case study, and we can see that using 5 traces in this PSPACE-complete scenario made it impossible to find a solution for some of the most demanding domains. Contrary to the previous case study, the difference between the syntactic and semantic metrics is larger here. Comparing the scores of both versions, we find that models that originally achieved mediocre scores when using the GTM as reference, are considerably sound and complete, reaching and overall score of 0.91 for both sem-Precision and sem-Recall.

>>>>

In this section we present a set of experiments designed to evaluate different aspects of \FAMA, from its performance to the quality of the learned models. Whenever applicable, we will consider both NP-complete and PSPACE-complete scenarios, in order to draw conclusions at both levels of complexity. We will also devote one experiment to assess the suitability of the evaluation metrics proposed in this paper.

\subsection{Setup}
We evaluate \FAMA on fourteen IPC domains that satisfy the \strips\ requirement~\cite{fox2003pddl2} and that are taken from the {\sc planning.domains} repository~\cite{muise2016planning}. Table \ref{tab:domain_features} compiles the features of these fourteen domains used in the experiments. From left to right, the columns report the name of the domain, number of actions, number of predicates, maximum arity of the actions, and maximum arity of the predicates. These are all features which affect the size of $P_\Lambda$ and hence, have an impact on the complexity of the learning task. %As can be seen in table \ref{tab:domain_features}, the selected domains range from, such as \emph{blocks} and \emph{visitall}, to complex ones, like \emph{floortile} and \emph{satellite}, with higher number of actions and predicates and with higher arity.

In the following, we explain the details of our experimental setup:

@@

\item {\bf Plan traces}. For each domain, we have generated 10 plan traces with 10 actions/intermediate states each using random walks. Depending on the experiment, these traces are used for training or testing, so more details will be provided later.

\item {\bf Planner}. The classical planner we use to solve the $P_\Lambda$ instances that result from our compilations is {\sc Madagascar}~\cite{rintanen2014madagascar}. We use {\sc Madagascar} for two reasons: 1) its ability to deal with planning instances populated with dead-ends, and 2) when the length of the plan traces is known (\FO action sequences or \FO state trajectories), the horizon of the solution plan is also known and can be provided to {\sc Madagascar}, addressing in this case NP-complete problems. This is because, SAT-based planners can apply the actions for programming preconditions in parallel during a single planning step (and the same for actions programming effects) because these actions do not interact. Hence, we know that the programming prefix of the solution plan can be solved in two steps, independently of the number of programming actions applied.

@@

\begin{table}[hbt!]
		\begin{center}
			\begin{tabular}{l|c|c|c|c|}	
				& \multicolumn{4}{c|}{Domain features}\\ \cline{2-5}
				 & {\bf \# actions} & {\bf \# predicates} & {\bf max action arity} & {\bf max predicate arity}  \\
				\hline
				Blocks & 4 & 5 & 2 & 2  \\
				Driverlog & 6 & 5 & 4 & 2  \\
				Ferry & 3 & 5 & 2 & 2  \\
				Floortile & 7 & 10 & 4 & 2  \\
				Grid & 5 & 9 & 4 & 2  \\
				Gripper & 3 & 4 & 3 & 2  \\
				Hanoi & 1 & 3 & 3 & 2  \\
				Miconic & 4 & 6 & 2 & 2  \\
				Npuzzle & 1 & 3 & 3 & 2  \\
				Parking & 4 & 5 & 3 & 2  \\
				Satellite & 5 & 8 & 4 & 2  \\
				Transport & 3 & 5 & 5 & 2  \\
				Visitall & 1 & 3 & 2 & 2  \\
				Zenotravel & 5 & 4 & 5 & 2
			\end{tabular}
		\end{center}
	\caption{\small Feature description of the 13 domains used in the experiments.}
	\label{tab:domain_features}	
\end{table}

\subsection{Impact of the size of the input knowledge}
This experiment evaluates how the size of the input knowledge, i.e., $\left|\mathcal{T}\right|$ affects the performance of \FAMA. The experiment analyzes the evolution of the computation time as well as the quality of the learned models as the size of $\mathcal{T}$ increases (from 1 to 10). Recall that each trace in $\mathcal{T}$ comprises 10 actions/intermediate states. To keep the experiment practicable, we introduce a 1000s timeout after which the learning process is killed and the learned models are assigned a score of 0 for both precision and recall. Our goals with this experiment are to:
\begin{enumerate}
	\item Identify the amount of input required by \FAMA to learn sound and complete models,
	\item Evaluate the scalability of \FAMA wrt the input size which, as stated in section \ref{sec:properties}, is one of the limiting factors of our approach.
\end{enumerate}

With this in mind we defined two case studies:
\begin{itemize}
	\item \textbf{\FO action sequence and \PO state trajectory}: This is the typical case addressed by most previous work, which corresponds to a NP-complete scenario. We are assuming a degree of observability of just 10\% for the state trajectory, meaning that each literal of a state has a 10\% chance of being observed.
	\item  \textbf{\NO action sequence and \NO state trajectory}: This is a PSPACE-complete scenario where both the action sequence and state trajectory are completely empty and only initial and final states are observed, i.e., $\tau = \tup{s_0, s_m}, \forall \tau \in \mathcal{T}$.
\end{itemize}

@@

With regards to the scalability of \FAMA, figure \ref{fig:np_time} paints a very interesting picture. In this figure, we can see an exponential increase in computation time for input sizes beyond 5 traces. Until an input size of 4 traces the computation time is below 1 s, but it reaches 153 s when the input is composed of 10 traces. These results match the performance of {\sc Madagascar}, since this planner is known to struggle with plan horizons beyond 150-200 steps (in our case 160 steps corresponds to 8 traces).

@@

The conclusion we can draw is that learning from few input samples is both our strong point and our limitation. Results on the NP-complete scenario show that the learned models are considerably sound and complete (to put these results into a richer perspective we will also compare ourselves with another approach in the following section). The PSPACE-complete scenario, on the other hand, shows that the learned models have less quality. Nevertheless, we argue that syntax-based metrics are not appropriate for the PSPACE-complete scenario due to the vast amount of reformulations that can appear in such under-constrained learning task and we will show proof of this in section \ref{minimal}.

\subsection{Comparison with \ARMS}
Here we compare the performance of \FAMA to that of \ARMS, one of the most well-known approaches learning planning models. \ARMS works under the assumption of plan traces with \FO action sequences and \NO state trajectories, so we will also adopt this assumption (\ARMS, as any other previous approaches for learning planning models cannot handle the PSPACE-complete scenarios).

We define a \emph{degree of observability} $\sigma$ for the state trajectory, ranging from 0\% to 100\%, that measures the probability of observing a literal, and evaluate both approaches as $\sigma$ increases using 5 plan traces as input. Note that when $\sigma = 0$ we have a \NO state trajectory, when $\sigma=100$ we have a \FO state trajectory and all cases in-between correspond to the \PO scenario. Figures \ref{fig:comparison_precision} and \ref{fig:comparison_recall} compare \FAMA and \ARMS in terms of precision and recall. The horizontal axis represents the degree of observability, while the vertical axis show the average precision (figure \ref{fig:comparison_precision}) or recall (figure \ref{fig:comparison_recall}) computed over the fourteen tested domains. Remarkably, figure \ref{fig:comparison_precision} shows that \FAMA dominates in terms of precision in all cases save for \FO state trajectories. Furthermore, in average, the models learned by our approach are 15\% to 24\% more precise than those learned by \ARMS. A similar conclusion can be drawn for recall (figure \ref{fig:comparison_recall}), where the difference is even larger, meaning that our models are more complete.

@@

@@

\FAMA outperforms \ARMS in this particular experiment, which is one with very limited input. By no means do these results imply that our approach is overall better for the NP-complete scenarios, just that it is able to learn better models when few plan traces are available. We are not reporting experiments where far more input traces are available, since \FAMA cannot handle so much input.

\subsection{Experiments with minimal input knowledge}
\label{minimal}
%In the previous experiments we have given a general view of the performance of \FAMA under different conditions.
So far, experiments have shown that \FAMA is able to learn with very small amounts of input knowledge, be it due to low observability or few training samples. In this section we want to take a closer look at the action models learned from minimal input knowledge. To that end, we will limit the input to only 2 plan traces and analyze the results under different levels of observability.

@@

Table \ref{tab:results_minimum_100_10} shows the results of the first case study. As was expected from the previous experiments, the recall scores are generally higher than the precision ones, and, in fact, the models learned for 6 of the tested domains were perfectly complete. Although precision is overall lower, it is interesting to notice that the soundness of most learned sets of positive effects is flawless. With regards to the computation time, we find times below the second in all cases except for one, "floor-tile", the most complex among the tested domains.

@@

The first thing to notice in table \ref{tab:results_minimum_30_30}, which collects the result of the PSPACE case study with 30\% observability, is that the values for some domains are missing. This is the case for "floor-tile", and "parking" which not only are fairly complex domains, but are also considered \emph{puzzle-like}, a quality that is known to put a strain on the planners. Interestingly enough, "hanoi" also qualifies as a \emph{puzzle-like} domain, and we can see how this is reflected on the computation time. With regards to the quality of the models, we find that the learned models retain a similar level of soundness, while their completeness has drastically worsened with respect to the previous case study. This is specially the case for the preconditions, which have gone from 0.94 recall to 0.54. %This results are difficult to explain, given that in this case study the intermediate states contain more information, but
Our intuition is that the input actions play a big role on how close the learned model are with respect to the GTM.

Next, we analyze the PSPACE case study with \NO action sequences and state trajectories (Table \ref{tab:results_minimum_0_0}). Contrary to what might be expected by looking at the previous table, we are able to find solutions for all domains. This is because, although the search is less guided, there are far more possible solutions for this learning task. This wider range of solutions is also evidenced by the quality of the learned models, which in spite of being compliant with the input data, are further from the original GTM. In the table we find that both precision and recall have seen their scores decreased to 0.6 and 0.5, respectively.

@@

We argue, however, that syntax-based metrics are not appropriate for scenarios with minimal observability as they cannot cope with the \emph{reformulations} that frequently occur in these circumstances. To illustrate this, Figure \ref{fig:macroaction} shows the PDDL encoding of the action model of the {\tt stack} operator learned from plan traces with \NO action sequences and state trajectories.

@@

This action model removes a block from on top of another block and puts it down on the table, all done in a single step. There are two main differences from the expected action model for {\tt stack}: 1) it is unstacking a block instead of stacking it, and 2) instead of the arm holding the block by the end of the action, the block ends on the table. We refer to the first difference as \emph{role swapping} and it happens when actions are removed from the trace. If no actions are present in the input traces, the name of the actions lose their utility and they are effectively anonymous actions, meaning that they can interchange their behavior with any comparable action model. The second difference is commonly known as \emph{macro-action} (in this case unstack + put-down) and it can happen when there are missing states in the input traces. Reformulated action models are perfectly valid and can be used to solve planning tasks. For instance, any blocks-world problem can be solved by unstacking all the blocks to the table (unstack + put-down) and then stacking them to meet the goal condition (pick-up + stack), so reformulated action models like the one in the figure can be used to solve common planning tasks. As we have shown, the \NO/\NO case study meets all the criteria needed for reformulations to happen, and this is the reason why scenarios such as this are better evaluated using semantic-based metrics.

@@

Next, we take a look at table \ref{tab:metric_comparison_0_0} that details the results of the \NO/\NO case study, and we can see that using 5 traces in this PSPACE-complete scenario made it impossible to find a solution for some of the most demanding domains. Contrary to the previous case study, the difference between the syntactic and semantic metrics is larger here. Comparing the scores of both versions, we find that models that originally achieved mediocre scores when using the GTM as reference, are considerably sound and complete, reaching and overall score of 0.91 for both sem-Precision and sem-Recall.

@@
